---
title: "Matching examples"
author: "Matthew Upson"
date: "9 June 2016"
output: html_document
---

```{r,warning=FALSE,message=FALSE}

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  error = FALSE
  )

```

```{r echo=FALSE,include=FALSE}
install.packages(
    c(
        "dplyr",
        "Matching",
        "magrittr",
        "cem"
    )
)
```

# Matching methods

## 1: Propensity score matching

The example shown here is reproduced from [Sekhon (2011)](http://sekhon.berkeley.edu/papers/MatchingJSS.pdf).
See also teh package homepage: <http://sekhon.berkeley.edu/matching/>.

```{r}

# Load required packages:

library(dplyr)
library(Matching)
library(cem)
library(magrittr)

```

This example uses the Lalonde dataset from the Matching package: 

```{r}
# Load the lalonde data 

data("lalonde") 

summary(lalonde)
str(lalonde)

```

```{r}
## outcome of interest and treatment indicator
## Re-label for later use

Y <- lalonde$re78
Tr <- lalonde$treat
  
# estimate first propensity model using a glm

```

The first substantice step is define a model for our glm. 
Note that this is a step that is peculiar to PSM compared to other matching methods, and is one reason why I believe it should not be favoured. 
Matching will only be as good as the fit of the model that is chosen, adding an additional level of complexity to the process.

```{r}

glm_formula <- formula(Tr ~ 
    age + educ + black + 
    hisp + married + nodegr + 
    re74 + re75
    )

## You can define the formula in the glm call istelf, but there are situations
## where it is preferable not to.


system.time(
glm1 <- glm(
  glm_formula,
  family = binomial,
  data = lalonde
)
)

```

We can check the outcome of the model as usual with:

```{r}

summary(glm1)

```

But in many respects, this is less important than checking the balance of the matching process itself, which we will do later

```{r}
## Now one to one matching with replacement. We will use the fitted values (the
## propensity score) from the glm as argument X within the Match call.

## These are values between 0 and 1. Just out of interest, we can plot these 
## out. This gives us an indication of how the propensity scores match across
## treatments.

data_frame(
  Tr = factor(Tr), 
  PSM = glm1$fitted.values
) %>%
  ggplot +
  aes(
    y = PSM,
    x = Tr,
    group = Tr,
    colour = Tr
  ) +
  geom_point()

```

Now run the Matching itself based on the above glm. 

```{r}

glm1$fitted %>% unname %>% head

system.time(
rr1 <- Match(
  Y = Y, 
  Tr = Tr, 
  # Take the fitted values 
  X = glm1$fitted.values,
  # See ?Match. Running on a DfE machien with a large dataset will probably
  # require ties to be broken with ties = FALSE. This means that a random match
  # will be chosen if multiple possibilities exist
  ties = TRUE,
  # See Sekhon 2011 for an explanation of ATT. In almost all cases, you will
  # want to use ATT.
  estimand = "ATT",
  # caliper sets the threshold to allow matching. If a potential match is beyond
  # the threshold caliper, then it is not included as a match. This can take 
  # either a scalar or a vector of scalars equal in length to the number of 
  # input varoables. In general I haven't tended to play with this, but it could
  # be used to improve matching outcomes. Note that the caliper itself is a
  # proportion of the standard deviation of the given variable. See ?mMtch
  caliper = 0.25
)
)

```

We need to call `summary()` to see the result of the match.
This will give us a sense of how succesful the matching has been.
We can see how many successful matches there have been.
By changing the `caliper` you will see that the number of succesful matches will change.
A higher `caliper` will give better matching, but a lower final sample for any statistical test.

```{r}

summary(rr1)

```

The `summary()` above also runs a statistical..

```{r}
## Now check the balance of our propensity score model. Note that this model uses many more variables than the model itself.

system.time(
MatchBalance(
  Tr ~ 
    age + I(age^2) + educ + 
    I(educ^2) + black + hisp +
    married + nodegr + re74 +
    I(re74^2) + re75 + I(re75^2) + 
    u74 + u75 + I(re74 * re75) + 
    I(age * nodegr) + I(educ * re74) + 
    I(educ * re75),
  match.out = rr1, 
  nboots = 1000, 
  data = lalonde
)
)

## Consider nodegr See summary statistics. First set shows means for treatment 
## and control and mean difference in standard deviation. Second set shows
## statistics based on the raw empirical-QQ plots, and hence are in the same
## scale as the variable being tested. The third set of statistics are the same
## empirical QQ statistics, but this time standardised. Finally there is the
## variance ration - the closer this is to 1, the better the match, and a
## p-value from a t-test of the difference of the means (not paired). For nodegr
## there were significant differences between treatments before and after
## matching.

```

## 2: Mahalanobis distance

```{r}
### Try matching again with Mahalanobis distance

X <- lalonde %>%
  dplyr::select(
    age,
    educ,
    black,
    hisp,
    married,
    nodegr,
    re74,
    re75
    )

```

Now one to one matching with replacement

```{r}

system.time(
rr2 <- Match(
  Y = Y, 
  Tr = Tr, 
  X = X,
  # Weight = 2 will use Mahalanobis distance as the distance metric
  Weight = 2
  )
)

```
Now look at the summary

```{r}

summary(rr2)

```

Now check match balance

```{r}

system.time(
MatchBalance(
  Tr ~ 
    age + I(age^2) + educ + 
    I(educ^2) + black + hisp +
    married + nodegr + re74 +
    I(re74^2) + re75 + I(re75^2) + 
    u74 + u75 + I(re74 * re75) + 
    I(age * nodegr) + I(educ * re74) + 
    I(educ * re75),
  match.out = rr2, 
  nboots = 1000, 
  data = lalonde
)
)

```

## 3: Genetic Matching

```{r}

X <- lalonde %>% 
  dplyr::select(
    age, educ, black, hisp, married, nodegr, re74, re75, u74, u75
  )

system.time(
BalanceMatrix <- lalonde %>%
  transmute(
    age, I(age^2), educ, I(educ^2), black, hisp,
    married, nodegr, re74, I(re74^2), re75, I(re75^2), 
    u74, u75, I(re74 * re75), I(age * nodegr), I(educ * re74),
    I(educ * re75)
  )
)

system.time(
gen1 <- GenMatch(
  Tr = Tr, 
  X = X, 
  BalanceMatrix = BalanceMatrix, 
  pop.size = 1000
)
)
```

check summary

```{r}

system.time(
mgen1 <- Match(
  Y = Y, 
  Tr = Tr, 
  X = X, 
  Weight.matrix = gen1
  )
)

system.time(
MatchBalance(
  Tr ~ 
    age + I(age^2) + educ + 
    I(educ^2) + black + hisp + 
    married + nodegr + re74 + 
    I(re74^2) + re75 + I(re75^2) + 
    u74 + u75 + I(re74 * re75) + 
    I(age * nodegr) + I(educ * re74) + I(educ * re75),
  data = lalonde, 
  match.out = mgen1, 
  nboots = 1000
)
)

summary(mgen1)

```

## Coarsened Exact Matching (CEM)

```{r}

X1 <- lalonde %>%
  dplyr::select(
    treat, age, educ, black, hisp, 
    married, nodegr, re74, re75, re78
  ) %>%
  mutate(
    black = factor(black),
    hisp = factor(hisp),
    married = factor(married),
    nodegr = factor(nodegr),
    re74,
    re75,
    re78
  )

imbalance(
  group = X1$treat, 
  data = dplyr::select(X1,-treat)
)

system.time(
mat <- cem(
  treatment = "treat",
  data = X1,
  drop = "re78",
  eval.imbalance = TRUE
)
)

mat

```


```{r}
system.time(
est <- att(mat, re78 ~ treat, data = X1) 
)
est


```

# Comparing methods on re74

Check before and after on an example variable. 
Ideally you would do this on all variables!

```{r, fig.width=12, fig.height=12}

par(mfrow = c(2,2))

qqplot(
  lalonde$re74[rr1$index.control], 
  lalonde$re74[rr1$index.treated],
  main = "Propensity score matching",
  xlab = "Control",
  ylab = "Treatment"
)

abline(
  coef = c(0, 1), 
  col = 2
)

qqplot(
  lalonde$re74[rr2$index.control], 
  lalonde$re74[rr2$index.treated],
  main = "Mahalanobis distance matching",
  xlab = "Control",
  ylab = "Treatment"
)

abline(
  coef = c(0, 1), 
  col = 2
)

qqplot(
  lalonde$re74[mgen1$index.control], 
  lalonde$re74[mgen1$index.treated],
  main = "Genetic matching algorithm",
  xlab = "Control",
  ylab = "Treatment"
)

abline(
  coef = c(0, 1), 
  col = 2
)

qqplot(
  X1 %>% dplyr::filter(mat$matched,treat == 0) %$% re74,
  X1 %>% dplyr::filter(mat$matched,treat == 1) %$% re74,
  main = "Coarsened Exact Matching",
  xlab = "Control",
  ylab = "Treatment"
)

abline(
  coef = c(0, 1), 
  col = 2
)

```

# Reproducibility info
```{r}
.libPaths()
sessionInfo()
```

# References

Sekhon, J. (2011). Multivariate and Propensity Score Matching Software with Automated Balance Optimization: The Matching Package for R. *Journal of statistical software*, 42, 7. Available at <http://sekhon.berkeley.edu/papers/MatchingJSS.pdf>

